Continuous Deductive Reasoning over Streamed RDF Data

### 2015-12-01 article.tex
many minor formatting edits and points of clarification. also slightly revised wording of conclusion to match introduction and new conclusions

@@

Based on this definition of streams, we go on to define a notion of
sliding time-based windows over streams, as well as states of those windows at
specific instants.  Every fact instantiated in the state of such a window at a
given instant is considered to be true at that instant.  We do not consider count-based
window functions, as the state of such windows (and therefore the truth of the
facts instantiated therein) is determined by the contents of the stream as a
whole, which is a form of aggregation that contrasts with that of time-based
windows: where the states of the latter are determined by some
function of each element of the stream (the time from the instant of its
appearance thereon), the states of the former are determined by a function of
the stream itself.  This is more similar to the aggregation functions of SQL,
such as $count()$, $max()$, etc., which are not part of the semantics of
positive Datalog or (outside of specific heavily restricted cases) OWL 2 RL
\citep{w3cowl2profiles}.

@@

We can reduce a WC-Datalog program $\Pi^{C,r} = \langle R, F^0,
\{ W^{S_x, r_x} \mid x \in (1,s) \} \rangle$
at an instant $i \in \mathbb{T}$ to its TA-Datalog equivalent by
applying the rules $R$ to the union of the persistent facts $F^0$ and all
finite states $W^{S_x,r_x}_i$ of windows $W^{S_x,r_x}$ over streams $S_x$ of range
$r_x \in \mathbb{N}$, for all $x \in (1,s)$; we refer to this as the {\em instantaneous state} of
the WC-Datalog program, written
$\Pi^{C,r}_i = \langle R, F^0 \cup \mathop{\cup}\limits_{x=1}^s W^{S_x,r_x}_i \rangle$.
As previously shown in \refdef{continuous datalog: ta datalog program}, this may then be further
reduced to the Datalog program
$\Pi = \langle R, strip(F^0 \cup \mathop{\cup}\limits_{x=1}^n W^{S_j,r_j}_i \rangle$.

@@

Thus far the translation from instantaneous sets of entailments to
streams of entailments has been stated axiomatically as the definition
of those streams.
However, it follows from these axioms that the instantaneous set
$E^T_{i}$ of instances entailed at any given instant $i$ by any single-window program
$\Pi^{C,r} = \langle R, F^0, \{ \langle S, r \rangle \} \rangle$ may be reconstructed from
windows of sufficient range over its entailed streams $P$ and $N$ at
$i$, as expressed in \refeqn{continuous datalog: entailment equivalence}.
{\nobreak\begin{equation} \labeleqn{continuous datalog: entailment equivalence}
  E^T_{i} = W^{P,x}_{i} \setminus W^{N,y}_{i}
\end{equation}
where ${y \geqslant x \geqslant r}$}. The minimum range $x$ is such to achieve
the \emph{complete} set of entailment instances $E^T_{i}$ of the
$\Pi^{C,r}_{i}$. In addition, the minimum range $y$ is such to achieve the
\emph{sound} set $E^T_{i}$.  It should be noted that the proof holds in the
case of $n$ windows of varying range, with $x \geqslant max(r_1,\dots,r_n)$.
We capture this derivation in the form of a \emph{valid window}, as per
\refdef{continuous datalog: valid window}.

@@

The decomposition of any stream into any number of more
sparse streams without loss of information allows us to alternatively define the
state of the valid window $W^{P \setminus N}$ at any instant $i \in \mathbb{T}$
as the union of the states of the set of sliding windows $\mathcal{W}^{P \setminus N}$
over each of the singleton streams $(\langle f, t \rangle \mapsto t^\prime)$
at $i$, the union of these streams being equal to $P$.  Remembering that
$P(\langle f, t \rangle) = t^\prime$ means that $\langle f, t \rangle$ occurs
in $P$ at $t^\prime$, the range of each sliding window in $\mathcal{W}^{P \setminus N}$
is the difference $d$ between $P(\langle f, t \rangle)$ (which equals $t$, as $P$ is
in-order) and $N(\langle f, t \rangle)$.  This difference is always positive by Axiom
\ref{axiom:continuous datalog: entailment precedes negation}, and bounded at the
upper limit $r$ of the program $\Pi^{C,r} \models \langle E^0, P, N \rangle$ by
\refdef{continuous datalog: CDP}.
\begin{gather}
  W^{P \setminus N}_i = \bigcup_{W^{p,d} \in \mathcal{W}^{P \setminus N}} W^{p,d}_i \\
  \mathcal{W}^{P \setminus N} = \left\{ W^{p,d} \,\middle\vert\, \begin{split}
    \langle f, t \rangle \in P \land \\
    p = (\langle f, t \rangle \mapsto t) \land \\
    d = N(\langle f, t \rangle) - t
  \end{split}\right\}
\end{gather}
Treated this way, the entailment of one WC-Datalog program, e.g. that in
\refeqn{continuous datalog: example WC-Datalog program}, may then be taken as a
part or all of the data in a successive WC-Datalog program, without modification
of the existing semantics or loss of information regarding the validity of the
entailments.
% However, in the interest
% of being concise, such a successive WC-Datalog program may express valid
% windows alongside any sliding windows.
For example, 
a program composed of the set of rules $R^\prime$, the set of base facts
$F^{0\prime}$, a sliding window of range $x \geqslant r$ over the
in-order stream $S^\prime$, and the entailment of
$\Pi^{C,r} \models \langle E^0, P, N \rangle$, would be expressed as in \refeqn{continuous datalog: successive WC-Datalog program}.
\begin{gather}
  \labeleqn{continuous datalog: example WC-Datalog program}
    \Pi^{C,r} \langle R, F^0, \{W^{S,r}\} \rangle \models \langle E^0, P, N \rangle \\
  \labeleqn{continuous datalog: successive WC-Datalog program}
    \Pi^{C,x} = \langle R^\prime, F^{0\prime} \cup E^0, \{ W^{S^\prime,x} \} \cup \mathcal{W}^{P \setminus N} \rangle \models \langle E^{0\prime}, P^\prime, N^\prime \rangle
\end{gather}

@@

  With regards to the notation of WC-Datalog, these built-in predicates
  may be represented by the set of facts $B$, or the set of persistent
  fact instances ${B^0 = \{ \langle b, 0 \rangle \mid b \in B \}}$, that
  are the collected extension of those predicates. A RIF-Core program may
  therefore be represented as a Datalog program
  ${\Pi = \langle R, F \cup B \rangle}$, a Time-Annotated RIF-Core program
  as a TA-Datalog program ${\Pi^{T} = \langle R, F^T \cup B^0 \rangle}$, a
  Continuous RIF-Core program as a C-Datalog program
  ${\Pi^{C} = \langle R, F^0 \cup B^0, S_1 \cup \dots \cup S_n \rangle}$
  and a Windowed Continuous RIF-Core program as a WC-Datalog program
  ${\Pi^{C,r} = \langle R, F^0 \cup B^0, \{W^{S_1,r_1}, \dots, W^{S_n,r_n}\} \rangle}$.
  Furthermore, ${\langle R, B \rangle \models \varnothing}$ and
  ${\langle R, B^0 \rangle \models \varnothing}$, for any set $R$ of
  \emph{safe} RIF-Core rules. As such, the form of entailment for each
  type of RIF-Core program is the same as that for the equivalent variety
  of Datalog program:
  \begin{align*}
    \Pi =&& \langle R, F \cup B \rangle &\models E \\
    \Pi^{T} =&& \langle R, F^T \cup B^0 \rangle &\models E^T \\
    \Pi^{C} =&& \langle R, F^0 \cup B^0, S \rangle &\models E^S \\
    \Pi^{C,r} =&& \langle R, F^0 \cup B^0, S, r \rangle &\models \langle E^0, P, N \rangle \\
    \Pi^{C,r}_i =&& \langle R, F^0 \cup B^0 \cup W^{S,r}_i \rangle &\models E^0 \cup E^T_i
  \end{align*}

@@

\begin{nestedsection}{CoDeR Model for Continuous Deductive Reasoning}{model}
  \emph{CoDeR} is an abstract model for \emph{Co}ntinuous \emph{De}ductive \emph{R}easoning over streamed RDF data that expresses the semantics of Windowed Continuous RIF-Core, thus supporting continuous reasoning with general knowledge expressed in both RIF-Core and OWL 2 RL.
  It is composed of a stream-based semantic data model and a minimal algebra of stream-to-stream operators, the former for expressing the persistent fact instances $F^0$ and $E^0$, the streams $S_i$, $P$ and $N$ and the window ranges $r_i$ for $i \in (1,s)$ of each ${\Pi^{C,r} = \langle R,F^0 \cup B^0, \{ W^{S_x,r_x} \mid x \in (1,s) \} \rangle \models \langle E^0,P,N \rangle}$, the latter for expressing the persistent set of rules $R$, the built-in predicates of RIF-Core $B^0$ and enforcing the window range $r$.
  It casts the problem of applying those rules to streamed RDF data as that of iterative pattern matching, in a manner inspired by the Rete pattern matching algorithm \citep{forgy79}, though it is meant to support any such algorithm rather than prescribing one itself.

  % Stream-based Data Model
  \begin{nestedsection}{Data Model}{model: data}
    The data model of CoDeR is composed of multiple-source multiple-sink data streams, as shown in \reffig{CoDeR: stream}.
  The data model of CoDeR is composed of multiple-source multiple-sink data streams, as shown in \reffig{CoDeR: stream}.
  These streams are based on the notion of in-order streams presented in \refdef{continuous datalog: stream}, but are divided into two classes: \emph{input}/\emph{output} streams and \emph{inter-operator} streams.
  In addition, the notion of an \emph{instance} is extended in \refdef{CoDeR: graph instance} from that of an instance of a \emph{fact} that became true at a specific instant, as presented in \refdef{continuous datalog: fact instance}.
  The streams of CoDeR are streams of this extended form of instance, as per \refdef{CoDeR: stream}, which are compatible with the ongoing work on continuous query semantics by the RSP working group\footnote{According to \url{https://github.com/streamreasoning/RSP-QL/blob/master/Semantics.md}, 4th Nov 2015, a CoDeR graph instance may be expressed as an RSP graph with two timestamps predicated as `became valid' and `ceased to be valid'.}.
  \begin{definition}[Graph Instance]\labeldef{CoDeR: graph instance}
    A \emph{graph instance} is an RDF graph with some additional temporal annotation regarding both the instant from which it became valid and the instant at which it will cease to be valid;
    it is a triple composed of the graph ${\{(s,p,o),\dots,(s',p',o')\}}$ and the pair of instants $e$ and $n$ that bound the interval ${(e,n]}$ for which the graph is valid, written ${\langle \{(s,p,o),\dots,(s',p',o')\},e,n \rangle}$.
    The names $e$ and $n$ are so named because they relate to the instants of the instance's {$e$ntailment} and {$n$egation} by the program.
    It should be noted that validity is a stronger concept than truth, as all facts within a graph that is valid in a system are understood to be true, but facts may be true despite not being represented in a graph that is valid within the system.
    Two graph instances containing the same facts and annotation for $e$ and $n$ are considered identical;
    two graph instances containing the same facts but different annotations for $e$ and/or $n$ are considered to be distinct.
  \end{definition}
  \begin{definition}[CoDeR Stream]\labeldef{CoDeR: stream}
    All \emph{streams} of CoDeR are sequences of graph instances, using the definition of \emph{sequence} used in \refdef{continuous datalog: stream} of streams in WC-Datalog.
    They are also \emph{in-order} with regards to their temporal annotation $e$, but potentially \emph{out-of-order} with regards to their temporal annotation $n$.
  \end{definition}
  \begin{definition}[Empty Stream]\labeldef{CoDeR: empty stream}
    The \emph{empty stream} is a CoDeR stream that contains no graph instances, and is denoted by the symbol $\varnothing$.
    This means that the stream will contain no graph instances that map to instant $i$, for all instants ${i \in \mathbb{T}}$.
    This stream may occur in specific expressions of any program, as described in \refsec{CoDeR: transformations}.
  \end{definition}

@@

  The second class of streams, inter-operator streams, are CoDeR streams that are both produced and consumed exclusively by the stream-to-stream operators of CoDeR's minimal algebra.
  These streams are sequences of the intermediate results of processing between steps in the production of entailments.
  As the production of entailments in CoDeR is cast as an iterative pattern matching problem, all intermediate results of this process will also take the form of instances of RDF graphs.
  However, unlike the graph instances in input and output streams, all instances in a given inter-operator stream will be of graphs that match the same sequenced graph pattern, as per \refdef{CoDeR: sequenced pattern}, yielding \refdef{CoDeR: inter-operator} for inter-operator streams.
  \begin{definition}[Sequential Conjunction]\labeldef{CoDeR: sequential conjunction}
    \emph{Sequential conjunction} is the restriction of conventional conjunction to express not only that two facts are true, but that one precedes the other, for some notion of weak precedence.
    In the context of CoDeR, where the validity of facts changes over time, the sequential conjunction of one fact $a$ with another $b$ is definable as in \refeqn{CoDeR: seq}.
    \begin{equation}
      \labeleqn{CoDeR: seq}
      a\ seq\ b \equiv
        \begin{split}
          a \in g_a \land \\
          b \in g_b \land \\
          \langle g_a, e_a, n_a \rangle \land \\
          \langle g_b, e_b, n_b \rangle \land \\
          e_a \leqslant e_b < min\left( n_a , n_b \right)
        \end{split}
    \end{equation}
    This means that not only are $a$ and $b$ valid at the same instant, but $a$ became valid during an instant prior to (or the same as) that at which $b$ became valid.
    For example, if we replace the rule ${c \Leftarrow a \land b}$ in the continuous program \reffig{CoDeR: consecutive states} with ${c \Leftarrow a\ seq\ b}$, then the entailment of each state (and so the streamed entailment of the program) would remain unchanged.
    However, if the rule that uses the complementary sequential conjunction ${c \Leftarrow b\ seq\ a}$ were substituted into that program, then its entailment would be empty, because $a$ strictly preceded $b$.
  \end{definition}
  \begin{definition}[Sequenced Pattern]\labeldef{CoDeR: sequenced pattern}
    A \emph{sequenced pattern} is a SPARQL advanced graph pattern \citep{w3csparql}, extended with the notion of ordering between the components of the graph pattern.
    Where SPARQL graph patterns may be composed using explicit disjunction (UNION and OPTIONAL clauses) or implicit conventional conjunction, sequenced graphs may also make use of explicit \emph{sequential conjunction} between its constituent components, as per \refdef{CoDeR: sequential conjunction}, where each component may be a triple pattern or another sequenced graph pattern.
    As such, the set of all sequenced graph patterns completely subsumes the set of advanced graph patterns of SPARQL, those being sequenced graph patterns that make no use of sequential conjunction.

@@

    The algebra fundamentally consists of the following set of stream-to-stream operators: ${\{{}_g\pi_t, \sigma_{triple}, \sigma_{external}, \coloneq_{head}, \cup, \rstreamjoin\}}$.
    These operators encapsulate the practical expression of the set of functions required to derive the entailment of WC-RIF-Core programs in the context of CoDeR input/output streams.
    These functions are those required for WC-Datalog programs, with the addition of two functions:
    a one-pass unary evaluation function $\sigma_{external}$ for the built-in predicates of RIF-Core;
    and a one-pass unary graph decomposition function ${{}_g\pi_t}$ to convert the CoDeR input stream of heterogeneous RDF graph instances into a stream of homogeneous RDF triple instances, the latter being the form of a stream $S_x$ of a program ${\Pi^{C,r} = \langle R, F^0, \{ W^{S_x, r_x} \mid x \in (1,n) \} \rangle}$.
    The continuous derivation of the entailment of WC-Datalog programs requires, in turn, the set of functions required for querying deductive databases using Datalog, with the addition of another two functions:
    a two-pass unary valid-window maintenance function $w$ to enforce the windows of range $r_x$, for $x \in (1,n)$, over the CoDeR input streams;
    and a two-pass unary window-to-stream function $s$ to extract the stream of novel graph instances from a sequence of consecutive valid-windows, closely related to the ${IStream()}$ function of CQL \citep{StanfordStream}.
    Finally, the functions required for querying deductive databases using Datalog are conventionally the complete set required of the relational algebra\footnote{This refers to the complete relational algebra less the functionality of set minus $\setminus$, which is analogous to the NaF operator $\sim$ that is not supported by positive Datalog.}, being ${\{\sigma,\pi,\Join,\cup\}}$ as described by \citet{codd70relationalModel}.
    However, three of these may be restricted when processing semantic data, as discussed in \refsec{background: semantic queries}, becoming:
    the one-pass unary equality-based triple-select $\sigma_{triple}$;
    the two-pass binary natural join $\Join^\ast$ between \emph{sets} of variable bindings for graph patterns (requiring valid-windows over streams as input, and producing valid-windows);
    and the one-pass unary template-project $\pi_{triple}$ that projects constant values and bindings for variables in graph patterns from graph instances into new triple instances.
    The one-pass binary set operator $\cup$ utilised by the relational algebra need only be modified to preserve the mapping of instances to instants when applied to CoDeR streams.

>>>>

It should be noted that a fact instance only indicates that the
fact is recognised as true at that instant.  This does \emph{not}
imply that the fact is true for \emph{only} that instant.  As such,
a fact instance denotes an interval of indefinite length from a
definite start for which the fact is to be considered true.

@@

Based on this definition of streams, we go on to define a notion of
sliding time-based windows over streams, as well as states of those windows at
specific instants.  Every fact instantiated in the state of such a window at a
given instant is considered to be true at that instant.  We do not consider count-based
window functions, as the state of such windows (and therefore the truth of the
facts instantiated therein) is determined by the contents of the stream as a
whole, which is a form of aggregation that contrasts with that of time-based
windows: where the states of the latter are determined by some
function of each element of the stream (the time from the instant of its
appearance thereon), the states of the former are determined by a function of
the stream itself.  This is more similar to the aggregation functions of SQL and ASP,
such as $count()$, $max()$, etc., which are not part of the semantics of
positive Datalog or (outside of specific heavily restricted cases) OWL 2 RL
\citep{w3cowl2profiles}.

@@

We can reduce a WC-Datalog program $\Pi^{C,r} = \langle R, F^0,
\{ W^{S_x, r_x} \mid x \in (1,s) \} \rangle$
at an instant $i \in \mathbb{T}$ to its TA-Datalog equivalent by
applying the rules $R$ to the union of the persistent facts $F^0$ and all
finite states $W^{S_x,r_x}_i$ of windows $W^{S_x,r_x}$ over streams $S_x$ of range
$r_x \in \mathbb{N}$, for all $x \in (1,s)$; we refer to this as the {\em instantaneous state} of
the WC-Datalog program, written
$\Pi^{C,r}_i = \langle R, F^0 \cup \mathop{\cup}\limits_{x=1}^s W^{S_x,r_x}_i \rangle$.
As previously shown in \refdef{continuous datalog: ta datalog program}, this may then be further
reduced to the Datalog program
$\Pi = \langle R, strip(F^0 \cup \mathop{\cup}\limits_{x=1}^n W^{S_x,r_x}_i \rangle$.

@@

Thus far the translation from instantaneous sets of entailments to
streams of entailments has been stated axiomatically as the definition
of those streams.
However, it follows from these axioms that the instantaneous set
$E^T_{i}$ of instances entailed at any given instant $i$ by any single-window program
$\Pi^{C,r} = \langle R, F^0, \{ W^{S, r} \} \rangle$ may be reconstructed from
windows of sufficient range over its entailed streams $P$ and $N$ at
$i$, as expressed in \refeqn{continuous datalog: entailment equivalence}.
{\nobreak\begin{equation} \labeleqn{continuous datalog: entailment equivalence}
  E^T_{i} = W^{P,x}_{i} \setminus W^{N,y}_{i}
\end{equation}
where ${y \geqslant x \geqslant r}$}. The minimum range $x$ is such to achieve
the \emph{complete} set of entailment instances $E^T_{i}$ of the
$\Pi^{C,r}_{i}$. In addition, the minimum range $y$ is such to achieve the
\emph{sound} set $E^T_{i}$.  It should be noted that the proof holds in the
case of $n$ windows of varying range, with $x \geqslant max(r_1,\dots,r_n)$.
We capture this derivation in the form of a \emph{valid window}, as per
\refdef{continuous datalog: valid window}.

@@

The decomposition of any stream into any number of more
sparse streams without loss of information allows us to alternatively define the
state of the valid window $W^{P \setminus N}$ at any instant $i \in \mathbb{T}$
as the union of the states of the set of sliding windows $\mathcal{W}^{P \setminus N}$
over each of the singleton streams $(\langle f, t \rangle \mapsto t^\prime)$
at $i$, the union of these streams being equal to $P$.  Remembering that
$P(\langle f, t \rangle) = t^\prime$ means that $\langle f, t \rangle$ occurs
in $P$ at $t^\prime$, the range of each sliding window in $\mathcal{W}^{P \setminus N}$
is the difference $d$ between $P(\langle f, t \rangle)$ (which equals $t$, as $P$ is
in-order) and $N(\langle f, t \rangle)$.  This difference is always positive by Axiom
\ref{axiom:continuous datalog: entailment precedes negation}, and bounded at the
upper limit $r$ of the program $\Pi^{C,r} \models \langle E^0, P, N \rangle$ by
\refdef{continuous datalog: CDP}.
\begin{gather}
  W^{P \setminus N}_i = \bigcup_{W^{p,d} \in \mathcal{W}^{P \setminus N}} W^{p,d}_i \\
  \mathcal{W}^{P \setminus N} = \left\{ W^{p,d} \,\middle\vert\, \begin{split}
    \langle f, t \rangle \in P \land \\
    p = (\langle f, t \rangle \mapsto t) \land \\
    d = N(\langle f, t \rangle) - t
  \end{split}\right\}
\end{gather}
Treated this way, the entailment of one WC-Datalog program, e.g. that in
\refeqn{continuous datalog: example WC-Datalog program}, may then be taken as a
part or all of the data in a successive WC-Datalog program, without modification
of the existing semantics or loss of information regarding the validity of the
entailments.
% However, in the interest
% of being concise, such a successive WC-Datalog program may express valid
% windows alongside any sliding windows.
For example, 
a program composed of the set of rules $R^\prime$, the set of base facts
$F^{0\prime}$, a sliding window of range $x \geqslant r$ over the
in-order stream $S^\prime$, and the entailment of
$\Pi^{C,r} \models \langle E^0, P, N \rangle$, would be expressed as in \refeqn{continuous datalog: successive WC-Datalog program}.
\begin{gather}
  \labeleqn{continuous datalog: example WC-Datalog program}
    \Pi^{C,r} \langle R, F^0, \{W^{S,r}\} \rangle \models \langle E^0, P, N \rangle \\
  \labeleqn{continuous datalog: successive WC-Datalog program}
    \begin{split}
      \Pi^{C,x} &= \langle R^\prime, F^{0\prime} \cup E^0, \{ W^{S^\prime,x} \} \cup \mathcal{W}^{P \setminus N} \rangle \\
      &\models \langle E^{0\prime}, P^\prime, N^\prime \rangle
    \end{split}
\end{gather}

@@

  With regards to the notation of WC-Datalog, these built-in predicates
  may be represented by the set of facts $B$, or the set of persistent
  fact instances ${B^0 = \{ \langle b, 0 \rangle \mid b \in B \}}$, that
  are the collected extension of those predicates. A RIF-Core program may
  therefore be represented as a Datalog program
  ${\Pi = \langle R, F \cup B \rangle}$, a Time-Annotated RIF-Core program
  as a TA-Datalog program ${\Pi^{T} = \langle R, F^T \cup B^0 \rangle}$, a
  Continuous RIF-Core program as a C-Datalog program
  ${\Pi^{C} = \langle R, F^0 \cup B^0, S_1 \cup \dots \cup S_n \rangle}$
  and a Windowed Continuous RIF-Core program as a WC-Datalog program
  ${\Pi^{C,r} = \langle R, F^0 \cup B^0, \{ W^{S_x,r_x} \mid x \in (1,n)\} \rangle}$.
  Furthermore, ${\langle R, B \rangle \models \varnothing}$ and
  ${\langle R, B^0 \rangle \models \varnothing}$, for any set $R$ of
  \emph{safe} RIF-Core rules. As such, the form of entailment for each
  type of RIF-Core program is the same as that for the equivalent variety
  of Datalog program:
  \begin{align*}
    \Pi =&& \langle R, F \cup B \rangle &\models E \\
    \Pi^{T} =&& \langle R, F^T \cup B^0 \rangle &\models E^T \\
    \Pi^{C} =&& \langle R, F^0 \cup B^0, S \rangle &\models E^S \\
    \Pi^{C,r} =&& \langle R, F^0 \cup B^0, S, r \rangle &\models \langle E^0, P, N \rangle \\
    \Pi^{C,r}_i =&& \langle R, F^0 \cup B^0 \cup W^{S,r}_i \rangle &\models E^0 \cup E^T_i
  \end{align*}

@@

\begin{nestedsection}{CoDeR Model for Continuous Deductive Reasoning}{model}
  \emph{CoDeR} is an abstract model for \emph{Co}ntinuous \emph{De}ductive \emph{R}easoning over streamed RDF data that expresses the semantics of Windowed Continuous RIF-Core, thus supporting continuous reasoning with general knowledge expressed in both RIF-Core and OWL 2 RL.
  It is composed of a stream-based semantic data model and a minimal algebra of stream-to-stream operators, the former for expressing the persistent fact instances $F^0$ and $E^0$, the streams $S_x$, $P$ and $N$ and the window ranges $r_x$ for $x \in (1,s)$ of each $\Pi^{C,r}$, the latter for expressing the persistent set of rules $R$, the built-in predicates of RIF-Core $B^0$ and enforcing the window ranges $r_x$.
  It casts the problem of applying those rules to streamed RDF data as that of iterative pattern matching, in a manner inspired by the Rete pattern matching algorithm \citep{forgy79}, though it is meant to support any such algorithm rather than prescribing one itself.

  % Stream-based Data Model
  \begin{nestedsection}{Data Model}{model: data}
    The data model of CoDeR is composed of multiple-source multiple-sink data streams, as shown in \reffig{CoDeR: stream}.
  The data model of CoDeR is composed of multiple-source multiple-sink data streams, as shown in \reffig{CoDeR: stream}.
  These streams are based on the notion of in-order streams presented in \refdef{continuous datalog: stream}, but are divided into two classes: \emph{input}/\emph{output} streams and \emph{inter-operator} streams.
  In addition, the notion of an \emph{instance} is extended in \refdef{CoDeR: graph instance} from that of an instance of a \emph{fact} that became true at a specific instant, as presented in \refdef{continuous datalog: fact instance}.
  The streams of CoDeR are streams of this extended form of instance, as per \refdef{CoDeR: stream}, which are compatible with the ongoing work on continuous query semantics by the RSP working group\footnote{According to \url{https://github.com/streamreasoning/RSP-QL/blob/master/Semantics.md}, 4th Nov 2015, a CoDeR graph instance may be expressed as an RSP graph with two timestamps predicated as `became valid' and `ceased to be valid'.}.
  \begin{definition}[Graph Instance]\labeldef{CoDeR: graph instance}
    A \emph{graph instance} is an RDF graph with some additional temporal annotation regarding both the instant from which it became valid and the instant at which it will cease to be valid;
    it is a triple composed of the graph ${\{(s,p,o),\dots,(s',p',o')\}}$ and the pair of instants $e$ and $n$ that bound the interval ${(e,n]}$ for which the graph is valid, written ${\langle \{(s,p,o),\dots,(s',p',o')\},e,n \rangle}$.
    The names $e$ and $n$ are so named because they relate to the instants of the instance's {$e$ntailment} and {$n$egation} by the program.
    It should be noted that validity is a stronger concept than truth, as all facts within a graph that is valid in a system are understood to be true, but facts may be true despite not being represented in a graph that is valid within the system.
    Two graph instances containing the same facts and annotation for $e$ and $n$ are considered identical;
    two graph instances containing the same facts but different annotations for $e$ and/or $n$ are considered to be distinct.
  \end{definition}
  \begin{definition}[CoDeR Stream]\labeldef{CoDeR: stream}
    All \emph{streams} of CoDeR are sequences of graph instances, using the definition of \emph{sequence} used in \refdef{continuous datalog: stream} of streams in WC-Datalog.
    They are also \emph{in-order} with regards to their temporal annotation $e$, but potentially \emph{out-of-order} with regards to their temporal annotation $n$.
    In practise, where a stream received from some source external to the model is not annotated with one or either, these instants are to be assigned by some trivial pre-processing;
    this will be based on either the existing annotation, if it exists, or the current time, and the sliding window definitions in the modelled $\Pi^{C,r}$.
  \end{definition}
  \begin{definition}[Empty Stream]\labeldef{CoDeR: empty stream}
    The \emph{empty stream} is a CoDeR stream that contains no graph instances, and is denoted by the symbol $\varnothing$.
    This means that the stream will contain no graph instances that map to instant $i$, for all instants ${i \in \mathbb{T}}$.
    This stream may occur in specific expressions of any program when the range of an operator is mutually exclusive to that of the operator that succeeds it in a dataflow.
  \end{definition}

@@

  Where many stream processing tasks (including continuous querying) are one-pass, the process of reasoning and inference is iterative/recursive.
  This theoretically necessitates the union of the output stream with the input stream ``upstream'' of the remaining operators of any processing plan in the form of a \emph{working stream}, as per \refdef{CoDeR: working stream}.
  \begin{definition}[Working Stream]\labeldef{CoDeR: working stream}
    The \emph{working stream} of a CoDeR-based system is a CoDeR stream that is the union of all the system's input streams and its output stream.
    It facilitates the \emph{iterative} application of the rules of the system to the streamed data, in the same manner as the \emph{working memory} of the Rete algorithm \citep{forgy79}.
  \end{definition}

  The second class of streams, inter-operator streams, are CoDeR streams that are both produced and consumed exclusively by the stream-to-stream operators of CoDeR's minimal algebra.
  These streams are sequences of the intermediate results of processing between steps in the production of entailments.
  As the production of entailments in CoDeR is cast as an iterative pattern matching problem, all intermediate results of this process will also take the form of instances of RDF graphs.
  However, unlike the graph instances in input and output streams, all instances in a given inter-operator stream will be of graphs that match the same sequenced graph pattern, as per \refdef{CoDeR: sequenced pattern}, yielding \refdef{CoDeR: inter-operator} for inter-operator streams.
  \begin{definition}[Sequential Conjunction]\labeldef{CoDeR: sequential conjunction}
    \emph{Sequential conjunction} is the restriction of conventional conjunction to express not only that two facts are true, but that one precedes the other, for some notion of weak precedence.
    In the context of CoDeR, where the validity of facts changes over time, the sequential conjunction of one fact $a$ with another $b$ is definable as in \refeqn{CoDeR: seq}.
    \begin{equation}
      \labeleqn{CoDeR: seq}
      a\ seq\ b \equiv
        \begin{split}
          a \in g_a & \land \\
          b \in g_b & \land \\
          \langle g_a, e_a, n_a \rangle & \land \\
          \langle g_b, e_b, n_b \rangle & \land \\
          e_a \leqslant e_b < min\left( n_a , n_b \right) &
        \end{split}
    \end{equation}
    This means that not only are $a$ and $b$ valid at the same instant, but $a$ became valid during an instant prior to (or the same as) that at which $b$ became valid.
    For example, if we replace the rule ${c \Leftarrow a \land b}$ in the continuous program \reffig{CoDeR: consecutive states} with ${c \Leftarrow a\ seq\ b}$, then the entailment of each state (and so the streamed entailment of the program) would remain unchanged.
    However, if the rule that uses the complementary sequential conjunction ${c \Leftarrow b\ seq\ a}$ were substituted into that program, then its entailment would be empty, because $a$ strictly preceded $b$.
  \end{definition}
  \begin{definition}[Sequenced Pattern]\labeldef{CoDeR: sequenced pattern}
    A \emph{sequenced pattern} is a SPARQL advanced graph pattern \citep{w3csparql}, extended with the notion of ordering between the components of the graph pattern.
    Where SPARQL graph patterns may be composed using explicit disjunction (UNION and OPTIONAL clauses) or implicit conventional conjunction, sequenced graphs may also make use of explicit \emph{sequential conjunction} between its constituent components, as per \refdef{CoDeR: sequential conjunction}, where each component may be a triple pattern or another sequenced graph pattern.
    As such, the set of all sequenced graph patterns completely subsumes the set of advanced graph patterns of SPARQL, those being sequenced graph patterns that make no use of sequential conjunction.

@@

    The algebra fundamentally consists of the following set of stream-to-stream operators: ${\{{}_g\pi_t, \sigma_{triple}, \sigma_{external}, \coloneq_{head}, \cup, \rstreamjoin\}}$.
    These operators encapsulate the practical expression of the set of functions required to derive the entailment of WC-RIF-Core programs in the context of CoDeR input/output streams.
    These functions are those required for WC-Datalog programs, with the addition of two functions:
    a one-pass unary evaluation function $\sigma_{external}$ for the built-in predicates of RIF-Core;
    and a one-pass unary graph decomposition function ${{}_g\pi_t}$ to convert the CoDeR input stream of heterogeneous RDF graph instances into a stream of homogeneous RDF triple instances, the latter being the form of a stream $S_x$ of a program ${\Pi^{C,r} = \langle R, F^0, \{ W^{S_x, r_x} \mid x \in (1,n) \} \rangle}$.
    The continuous derivation of the entailment of WC-Datalog programs requires, in turn, the set of functions required for querying deductive databases using Datalog, with the addition of another two functions:
    a two-pass unary valid-window maintenance function $w$ to enforce the windows of range $r_x$, for $x \in (1,n)$, over the CoDeR input streams;
    and a two-pass unary window-to-stream function $s$ to extract the stream of novel graph instances from a sequence of consecutive valid-windows, closely related to the ${IStream()}$ function of CQL \citep{StanfordStream}.
    Finally, the functions required for querying deductive databases using Datalog are conventionally the complete set required of the relational algebra\footnote{This refers to the complete relational algebra less the functionality of set minus $\setminus$, which is analogous to the NaF operator $\sim$ that is not supported by positive Datalog.}, being ${\{\sigma,\pi,\Join,\cup\}}$ as described by \citet{codd70relationalModel}.
    However, three of these may be restricted when processing semantic data, becoming:
    the one-pass unary equality-based triple-select $\sigma_{triple}$;
    the two-pass binary natural join $\Join^\ast$ between \emph{sets} of variable bindings for graph patterns (requiring valid-windows over streams as input, and producing valid-windows);
    and the one-pass unary template-project $\pi_{triple}$ that projects constant values and bindings for variables in graph patterns from graph instances into new triple instances.
    The one-pass binary set operator $\cup$ utilised by the relational algebra need only be modified to preserve the mapping of instances to instants when applied to CoDeR streams.

    All one-pass functions required to derive the entailment of a WC-RIF-Core program expressed by an equivalent stream-to-stream operator in CoDeR, as detailed in \reftab{WC-RIF-Core-functions-to-CoDeR-operators}.
    \begin{table*}
        \centering
        \begin{tabular}{r || c | c | c | c | c | c | c | c}
          WC-RIF-Core & $\sigma_{triple}$ & $\sigma_{external}$ & ${}_g\pi_t$ & $\pi_{triple}$ & $\cup$ & $w$ & $s$ & $\Join^\ast$ \\ \hline
          CoDeR & $\sigma_{triple}$ & $\sigma_{external}$ & ${}_g\pi_t$ & $\coloneq_{head}$ & $\cup$ & \multicolumn{3}{c}{$\rstreamjoin \cup \rstreamjoin$}
        \end{tabular}
        \caption[WC-RIF-Core Functions to CoDeR Operators]{The mapping of each function of WC-RIF-Core to their expression using the operators of the CoDeR algebra.}
        \labeltab{WC-RIF-Core-functions-to-CoDeR-operators}
      \end{table*}
      The two-pass functions $\{w,s,\Join^\ast\}$, as shown in the same table, are combined into a single stream-to-stream operator in CoDeR in order to make the algebra universally stream-to-stream in nature.
      This is possible because $w$ is solely used to produce a set from a stream, $s$ to produce a stream from a set, and $\Join^\ast$ is the only function that must specifically act between sets;
      it follows that the former two will always, and need only, be used in direct conjunction with the latter in the form $s(w(X) \Join^\ast w(Y))$, where $X$ and $Y$ are inter-operator streams.
